---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

devtools::load_all(quiet = TRUE)
```

# saeproj.multilevel

## Author

Nazlya Rahma Susanto, Azka Ubaidillah

## Maintainer

Nazlya Rahma Susanto [susantonazlya@gmail.com](mailto:susantonazlya@gmail.com)

## Description

The **saeproj.multilevel** package provides tools for *Small Area Estimation* (SAE) using a projection estimator with a multilevel regression model.

The method is designed for two-survey settings:

- a smaller model survey, `data_model`, that contains the response variable and auxiliary predictors;
- a larger projection survey, `data_proj`, that contains auxiliary predictors and survey design information, but does not contain the response variable.

The main function is:

```r
sae_ml_linear()
```

The function fits a linear multilevel regression model using `lme4::lmer()`, generates unit-level predictions for the projection dataset, aggregates those predictions by domain using survey design information, and applies a design-based residual correction.

The final projection estimator is:

```r
estimate_final = estimate_synthetic + correction
```

The plug-in variance is calculated as:

```r
variance_final = variance_synthetic + variance_correction
```

The synthetic projection component and residual correction component are stored in:

```r
result$estimation_details
```

## Installation

The development version of `saeproj.multilevel` can be installed from GitHub with:

```r
# install.packages("devtools")
devtools::install_github("rahmanazlya02/saeproj.multilevel")
```

To install the package together with the vignette, use:

```r
# install.packages("devtools")
devtools::install_github(
  "rahmanazlya02/saeproj.multilevel",
  build_vignettes = TRUE,
  dependencies = TRUE
)
```

After installation, the vignette can be opened with:

```r
browseVignettes("saeproj.multilevel")
```

Or directly:

```r
vignette(
  "sae_ml_linear",
  package = "saeproj.multilevel"
)
```

## Dependencies

The package imports:

- `lme4` — for fitting linear multilevel regression models;
- `survey` — for survey design and domain-level aggregation;
- `dplyr` — for joining estimation components;
- `cli` — for errors and selected warnings;
- `reformulas` — for parsing multilevel model formulas.

## Package datasets

The package includes two simulated datasets generated from one fixed replication of the study-simulation design.

```{r load-data}
data("saeml_modelsvy")
data("saeml_projsvy")
```

### `saeml_modelsvy`

`saeml_modelsvy` is a small model-survey dataset containing:

- 250 observations;
- 50 domains identified by `kab_kota`;
- 5 sampled units in each domain;
- the target variable `Y`;
- unit-level auxiliary variables `X1`, `X2`, `X3`, and `X4`;
- area-level auxiliary variables `Z1` and `Z2`;
- the survey weight variable `WEIND`;
- no separate PSU or cluster variable.

### `saeml_projsvy`

`saeml_projsvy` is a large projection-survey dataset containing:

- 15,000 observations;
- 50 domains identified by `kab_kota`;
- 300 sampled units in each domain;
- unit-level auxiliary variables `X1`, `X2`, `X3`, and `X4`;
- area-level auxiliary variables `Z1` and `Z2`;
- the survey weight variable `WEIND`;
- no target variable `Y`;
- no separate PSU or cluster variable.

The two datasets are drawn from the same simulated population and do not contain overlapping sampled units.

```{r data-dimensions}
dim(saeml_modelsvy)

dim(saeml_projsvy)
```

## Example

### Fit the multilevel projection estimator

```{r example-model}
result <- sae_ml_linear(
  formula = Y ~ X1 + X2 + X3 + X4 + Z1 + Z2 + (1 | kab_kota),
  data_model = saeml_modelsvy,
  data_proj = saeml_projsvy,
  domain = "kab_kota",
  cluster_ids = ~1,
  weight = "WEIND",
  strata = "kab_kota",
  summary_function = "mean"
)

result
```

The package datasets do not contain a separate PSU or cluster variable. Therefore, the example uses:

```r
cluster_ids = ~1
```

This specifies an unclustered survey-design structure. The variable `id_individu` is only a unique sampled-unit identifier and is not used as a PSU or cluster identifier.

### Domain-level estimates

The final domain-level estimates are stored in:

```r
result$estimates
```

The complete results for all 50 domains are shown below.

```{r final-estimates}
result$estimates
```

The output contains:

| Column | Description |
|---|---|
| domain variable(s) | Domain identifier column(s), based on the `domain` argument |
| `estimate` | Final projection estimate with design-based residual correction |
| `variance` | Plug-in variance of the final estimate |
| `se` | Standard error, computed as `sqrt(variance)` |
| `rse` | Relative standard error in percent |

The same result can be extracted for further analysis with:

```r
as.data.frame(result)
```

## Estimation components

Detailed estimation components are stored in:

```r
result$estimation_details
```

The complete synthetic estimate, residual correction, final estimate, variance, and sample-size information for all 50 domains are shown below.

```{r estimation-components}
result$estimation_details
```

This table contains:

| Column | Description |
|---|---|
| domain variable(s) | Domain identifier column(s) |
| `estimate_synthetic` | Synthetic projection estimate |
| `variance_synthetic` | Variance of the synthetic projection estimate |
| `correction` | Design-based residual correction |
| `variance_correction` | Variance of the residual correction |
| `estimate_final` | Final estimate, computed as `estimate_synthetic + correction` |
| `variance_final` | Final variance, computed as `variance_synthetic + variance_correction` |
| `se_final` | Standard error of the final estimate |
| `rse_final` | Relative standard error of the final estimate |
| `n_model` | Number of observations in the domain in `data_model` |
| `n_proj` | Number of observations in the domain in `data_proj` |

### Synthetic projection component

The function returns the projection estimator with a design-based residual correction by default.

The complete synthetic component for all 50 domains is available below.

```{r synthetic-component}
result$estimation_details[, c(
  "kab_kota",
  "estimate_synthetic",
  "variance_synthetic"
)]
```

## Direct estimator

Set `return_direct = TRUE` to return direct design-based estimates from `data_model`.

```r
result_direct <- sae_ml_linear(
  formula = Y ~ X1 + X2 + X3 + X4 + Z1 + Z2 + (1 | kab_kota),
  data_model = saeml_modelsvy,
  data_proj = saeml_projsvy,
  domain = "kab_kota",
  cluster_ids = ~1,
  weight = "WEIND",
  strata = "kab_kota",
  summary_function = "mean",
  return_direct = TRUE
)

result_direct$direct_estimator
```

The direct estimator is stored separately and does not replace the projection estimator.

## Print and summary

A concise output can be displayed with:

```{r print-result}
print(result)
```

A compact summary can be displayed with:

```{r summary-result}
summary(result)
```

The `summary()` method displays the formula, estimator type, number of domains, selected model diagnostics, and a preview of the final estimates.

Full model output can be accessed from the fitted `lmerMod` object:

```{r fitted-model}
fit <- result$fitted_model

summary(fit)
```

## Retaining unit-level predictions

Set `keep_unit = TRUE` to store unit-level projection data and model residual data.

```r
result_ku <- sae_ml_linear(
  formula = Y ~ X1 + X2 + X3 + X4 + Z1 + Z2 + (1 | kab_kota),
  data_model = saeml_modelsvy,
  data_proj = saeml_projsvy,
  domain = "kab_kota",
  cluster_ids = ~1,
  weight = "WEIND",
  strata = "kab_kota",
  summary_function = "mean",
  keep_unit = TRUE
)

head(result_ku$unit_projection)
head(result_ku$unit_model_residual)
```

When `keep_unit = TRUE`:

- `result_ku$unit_projection` contains `data_proj` with the unit-level prediction column `.prediction`;
- `result_ku$unit_model_residual` contains `data_model` with `.fitted_model` and `.model_residual`.

## Model diagnostics

Model diagnostics are stored in:

```r
result$diagnostics
```

```{r diagnostics}
data.frame(
  icc = result$diagnostics$icc,
  singular_fit = result$diagnostics$singular_fit,
  convergence = result$diagnostics$convergence,
  sigma = result$diagnostics$sigma,
  residual_variance = result$diagnostics$residual_variance,
  REML = result$diagnostics$REML,
  AIC = result$diagnostics$AIC,
  BIC = result$diagnostics$BIC
)
```

The estimated random effects for all domain groups can be inspected directly:

```{r random-effects}
lme4::ranef(result$fitted_model)$kab_kota
```

Residual diagnostics can be inspected from the fitted model:

```r
fit <- result$fitted_model

plot(
  fitted(fit),
  resid(fit),
  xlab = "Fitted values",
  ylab = "Residuals",
  main = "Residuals vs Fitted"
)

abline(h = 0, lty = 2)

qqnorm(resid(fit))
qqline(resid(fit))
```

## Model parameters

Estimated model parameters are stored in:

```r
result$model_parameters
```

```{r model-parameters}
result$model_parameters$fixed_effects

result$model_parameters$variance_components

result$model_parameters$residual_variance
```

## Notes

Run-specific notes are stored in:

```{r model-notes}
result$notes
```

The notes are intentionally concise and are not printed automatically by `summary()`.

They may include information such as:

- removed zero-variance predictors;
- singular model fit;
- convergence issues;
- random-slope or complex random-effect structures where a simple ICC is not computed;
- out-of-sample domains with zero residual correction;
- negative plug-in variance clamped to zero.

Out-of-sample domains are not treated as warnings because they are expected in SAE projection. They are recorded in `result$notes`.

## Multiple domain variables

The `domain` argument accepts a character scalar, a character vector, or a one-sided formula.

The following example uses both `prov` and `kab_kota` as domain identifiers.

```r
result_multi <- sae_ml_linear(
  formula = Y ~ X1 + X2 + X3 + X4 + Z1 + Z2 + (1 | kab_kota),
  data_model = saeml_modelsvy,
  data_proj = saeml_projsvy,
  domain = c("prov", "kab_kota"),
  cluster_ids = ~1,
  weight = "WEIND",
  strata = "kab_kota",
  summary_function = "mean"
)

result_multi$estimates
```

## Survey design specification

The arguments `cluster_ids`, `weight`, and `strata` are used in the aggregation step through `survey::svydesign()`.

### Simulated package data

The simulated datasets included in the package do not contain a separate PSU or cluster variable. Therefore, the package examples use:

```r
cluster_ids = ~1
weight = "WEIND"
strata = "kab_kota"
```

Here, `cluster_ids = ~1` specifies an unclustered survey-design structure.

### Survey data with PSU clustering

For a real survey with a PSU or cluster variable, provide the actual PSU identifier in `cluster_ids`.

The following code is illustrative. Replace `psu_id`, `survey_weight`, and `stratum` with the corresponding variable names in your data.

```r
result_clustered <- sae_ml_linear(
  formula = Y ~ X1 + X2 + X3 + X4 + Z1 + Z2 + (1 | kab_kota),
  data_model = data_model,
  data_proj = data_proj,
  domain = "kab_kota",
  cluster_ids = "psu_id",
  weight = "survey_weight",
  strata = "stratum",
  summary_function = "mean",
  nest = TRUE
)
```

In this specification:

- `psu_id` identifies the primary sampling unit or cluster;
- `survey_weight` identifies the sampling weight;
- `stratum` identifies the sampling stratum;
- `nest = TRUE` indicates that PSUs are nested within strata.

Use `cluster_ids = ~1` when the survey design does not include a separate PSU or cluster variable.

## Output object structure

`sae_ml_linear()` returns an S3 object of class `"sae_ml_linear"`.

Typical components are:

| Component | Description |
|---|---|
| `$call` | The matched function call |
| `$formula` | The model formula used after preprocessing |
| `$estimator` | Estimator type; currently always `"bias_corrected"` |
| `$fitted_model` | The fitted `lmerMod` object from `lme4::lmer()` |
| `$model_parameters` | Fixed effects, random effects, variance components, residual SD, and residual variance |
| `$estimates` | Final domain-level estimates |
| `$estimation_details` | Synthetic estimate, correction, final estimate, and sample sizes per domain |
| `$diagnostics` | Model diagnostics: ICC when applicable, random-effect structure, singular fit, convergence, sigma, residual variance, REML, logLik, AIC, and BIC |
| `$notes` | Concise run-specific notes |
| `$unit_projection` | Unit-level `data_proj` with `.prediction`, only if `keep_unit = TRUE` |
| `$unit_model_residual` | Unit-level `data_model` with `.fitted_model` and `.model_residual`, only if `keep_unit = TRUE` |
| `$direct_estimator` | Direct design-based estimates, only if `return_direct = TRUE` |

## S3 methods

| Method | Behaviour |
|---|---|
| `print(result)` | Prints formula, estimator, number of domains, and a preview of `$estimates` |
| `summary(result)` | Prints selected diagnostics and a preview of final estimates |
| `as.data.frame(result)` | Returns `result$estimates` |

## Function interface

```r
sae_ml_linear(
  formula,
  data_model,
  data_proj,
  domain,
  cluster_ids = ~1,
  weight = NULL,
  strata = NULL,
  summary_function = "mean",
  keep_unit = FALSE,
  seed = 1,
  control = lme4::lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 2e5)
  ),
  return_direct = FALSE,
  ...
)
```

| Argument | Description |
|---|---|
| `formula` | `lme4::lmer()`-style formula containing at least one random-effect term |
| `data_model` | Model survey data frame containing the response, predictors, grouping variables, domain variable(s), and survey design variables |
| `data_proj` | Projection survey data frame containing predictors, grouping variables, domain variable(s), and survey design variables; the response is not required |
| `domain` | Domain variable name(s): character scalar, character vector, or one-sided formula |
| `cluster_ids` | PSU or cluster variable for survey design; use `~1` for no clustering |
| `weight` | Survey weight variable; use `NULL` for equal weights |
| `strata` | Stratification variable; use `NULL` if not applicable |
| `summary_function` | Domain-level statistic: `"mean"` or `"total"` |
| `keep_unit` | If `TRUE`, unit-level predictions and residuals are stored in the output |
| `seed` | Integer seed used before model fitting |
| `control` | `lme4::lmerControl()` object passed to `lme4::lmer()` |
| `return_direct` | If `TRUE`, direct design-based estimates from `data_model` are returned |
| `...` | Additional named arguments passed to `survey::svydesign()`, for example `nest = TRUE` |

The `weight` argument identifies the survey weight column used in both `data_model` and `data_proj`. The column name must be the same in both datasets, but the weight values may differ.

In `data_model`, weights are used for residual correction and optional direct estimation. In `data_proj`, weights are used for synthetic projection aggregation.

## Methodological notes

- The working model is a linear multilevel regression model fitted with `lme4::lmer()` using restricted maximum likelihood estimation.
- The user fully specifies the fixed-effect and random-effect structure through the `formula` argument.
- Prediction uses `re.form = NULL` and `allow.new.levels = TRUE`.
- For grouping levels observed in `data_model`, predictions include the estimated random-effect contribution.
- For grouping levels appearing only in `data_proj`, the random-effect contribution is set to zero, so prediction uses the fixed part of the model.
- ICC is a diagnostic measure and does not determine the prediction rule.
- A simple ICC is computed only for pure random-intercept structures.
- Fixed-effect categorical predictors in `data_proj` must not contain levels that are absent from `data_model`.
- Fixed-effect predictors with zero variance in `data_model` are removed automatically before model fitting.
- Survey design arguments (`cluster_ids`, `weight`, and `strata`) are used in the aggregation step through `survey::svydesign()` and `survey::svyby()`.
- The plug-in variance is approximate and does not fully account for uncertainty in estimated multilevel model parameters.
- Missing values are not removed automatically. The function stops with an informative error if required variables contain missing values.

## Summary function

- The argument `summary_function` supports `"mean"` and `"total"` because both are linear domain parameters.
- For `"mean"`, the synthetic component and residual correction are aggregated using `survey::svymean`.
- For `"total"`, both components are aggregated using `survey::svytotal`, so the estimate and variance are returned on the total scale.
- The `"total"` option should only be used when survey weights are appropriate expansion weights for population totals.

## References

Bates, D., Maechler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software, 67*(1), 1–48.

Finch, W. H., Bolin, J. E., & Kelley, K. (2014). *Multilevel Modeling Using R*. CRC Press.

Food and Agriculture Organization of the United Nations. (2021). *Guidelines on Data Disaggregation for SDG Indicators Using Survey Data* (1st ed.). https://doi.org/10.4060/cb3253en

Hox, J. J., Moerbeek, M., & van de Schoot, R. (2018). *Multilevel Analysis: Techniques and Applications* (3rd ed.). Routledge.

Kim, J. K., & Rao, J. N. K. (2012). Combining data from two independent surveys: A model-assisted approach. *Biometrika, 99*(1), 85–100.

Moura, F. A. S., & Holt, D. (1999). Small area estimation using multilevel models. *Survey Methodology, 25*(1), 73–80.

Rao, J. N. K., & Molina, I. (2015). *Small Area Estimation* (2nd ed.). Wiley.
